The CrypTool Script: \\ Cryptography, Mathematics and more \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free eLearning program CrypTool\\ (8th edition -- distributed with CrypTool version 1.4.10)} \vspace{60pt}

### 2007-07-05 script/en/chapters/moderncryptography.tex
Update scripts: minimal correction (1.4.10 in menutree)

Let $ p $ be a prime number and let $ g \in \Z_p^\ast=\{0,1,\ldots,p-1\} $. Then 
the discrete exponential function base $ g $  is defined as
\[ e_g : k \longrightarrow y:=g^k \mod p, \quad 1\leq k \leq p-1. \]
The inverse function is called a discrete logarithm function $ \log_g $; the 
following holds:
\[ \log_g (g^k) =k. \]
\index{Exponential function!discrete} The problem of the discrete logarithm (in 
$ \Z_p^\ast$) is understood to be as follows:
\[ {\rm Given~ } p,g {\rm ~and~ } y, {\rm~determine~} k {\rm ~such~that~ } 
y=g^k \mod p {\rm .}\]
It is much more difficult to calculate the discrete logarithm than to evaluate 
the discrete exponential function (see chapter \ref{MultOrdPrimitveRoot}).
There are several procedures for calculating the discrete logarithm 
\cite{4Stinson1995}\index{Stinson 1995}:
\vskip + 5pt
\begin{center}
\begin{tabular}{|l|l|}\hline
Name                 &        Complexity \\ \hline \hline
Babystep-Giantstep &         $ O(\sqrt{p}) $ \\ \hline
Silver-Pohlig-Hellman    &    polynomial in $ q, $ the greatest \\
&  prime factor of $ p-1. $ \\ \hline
Index-Calculus &             $ O(e^{(1+o(1)) \sqrt{\ln (p) \ln (\ln (p))}}) $ \\ 
\hline
\end{tabular}
\end{center}
\vskip +5 pt
\index{Silver} \index{Pohlig, S. C.} \index{Hellman, Martin}
\index{Baby-Step-Giant-Step}

@@

\paragraph{Calculating the discrete logarithm }
A simple procedure for calculating the discrete logarithm of a group element, 
that is considerably more efficient than simply trying all possible values for $ 
k, $ is the \index{Baby-Step-Giant-Step} Baby-Step-Giant-Step algorithm.
\begin{theorem}\label{thm-cry-bsgs}[Baby-Step-Giant-Step algorithm]
Let $ G $ be a group and $ g \in G. $ Let $ n $ be the smallest natural number 
with $ |G|\leq n^2. $ Then the discrete logarithm of an element $ h 
\in G $ can be calculated base $ g $ by generating two lists each containing $ n 
$ elements and comparing these lists.\\
In order to calculate these lists, we need $ 2n $ group operations.
\end{theorem}

\begin{Proof}{}
  First create the two lists \\
Giant-Step list: $ \{1,g^n,g^{2n}, \ldots, g^{n \cdot n}\}, $\\
Baby-Step list: $ \{ hg^{-1} , hg^{-2} , \ldots , hg^{-n} \}. $ \par
If $ g^{jn} = hg^{-i}, $ i.e. $ h = g^{i+jn}, $ then the problem is solved. If 
the lists are disjoint, then $ h $ cannot be represented as $ g^{i + jn}, i, 
j\leq n,$. As all powers of $ g $ are thus recorded, the logarithm problem does 
not have a solution.
\end{Proof}

You can use the Baby-Step-Giant-Step algorithm to demonstrate that it is much 
more difficult to calculate the discrete logarithm than to calculate the 
discrete exponential function. If the numbers that occur have approximately 1000 
bits in length, then you only need around 2000 multiplications to calculate $ 
g^k $ but around $ 2^{500} \approx 10^{150} $ operations to calculate the 
discrete logarithm using the Baby-Step-Giant-Step algorithm. \\
In addition to the Baby-Step-Giant-Step algorithm, there are also numerous other 
procedures for calculating the discrete logarithm \cite{4Stinson1995}\index{Stinson 1995}.

>>>>

Let $ p $ be a prime number and let $ g \in \Z_p^\ast=\{0,1,\ldots,p-1\} $. Then 
the discrete exponential function base $ g $  is defined as
\[ e_g : k \longrightarrow y:=g^k \mod p, \quad 1\leq k \leq p-1. \]
The inverse function is called a discrete logarithm function $ \log_g $; the 
following holds:
\[ \log_g (g^k) =k. \]
\index{Exponential function!discrete} The problem of the discrete logarithm (in 
$ \Z_p^\ast$) is understood to be as follows:
\[ {\rm Given~ } p,g {\rm ~and~ } y, {\rm~determine~} k {\rm ~such~that~ } 
y=g^k \mod p {\rm .}\]
It is much more difficult to calculate the discrete logarithm than to evaluate 
the discrete exponential function (see chapter \ref{MultOrdPrimitveRoot}).
There are several procedures for calculating the discrete logarithm 
\cite{4Stinson1995}\index{Stinson 1995}:
\vskip + 5pt
\begin{center}
\begin{tabular}{|l|l|}\hline
Name                 &        Complexity \\ \hline \hline
Baby-step-giant-step &         $ O(\sqrt{p}) $ \\ \hline
Silver-Pohlig-Hellman    &    polynomial in $ q, $ the greatest \\
&  prime factor of $ p-1. $ \\ \hline
Index-Calculus &             $ O(e^{(1+o(1)) \sqrt{\ln (p) \ln (\ln (p))}}) $ \\ 
\hline
\end{tabular}
\end{center}
\vskip +5 pt
\index{Silver} \index{Pohlig, S. C.} \index{Hellman, Martin}
\index{Baby-step-giant-step}

@@

\paragraph{Calculating the discrete logarithm }
A simple procedure for calculating the discrete logarithm of a group element, 
that is considerably more efficient than simply trying all possible values for $ 
k, $ is the \index{Baby-step-giant-step} baby-step-giant-step algorithm.
\begin{theorem}\label{thm-cry-bsgs}[baby-step-giant-step algorithm]
Let $ G $ be a group and $ g \in G. $ Let $ n $ be the smallest natural number 
with $ |G|\leq n^2. $ Then the discrete logarithm of an element $ h 
\in G $ can be calculated base $ g $ by generating two lists each containing $ n 
$ elements and comparing these lists.\\
In order to calculate these lists, we need $ 2n $ group operations.
\end{theorem}

\begin{Proof}{}
  First create the two lists \\
Giant-step list: $ \{1,g^n,g^{2n}, \ldots, g^{n \cdot n}\}, $\\
Baby-step list: $ \{ hg^{-1} , hg^{-2} , \ldots , hg^{-n} \}. $ \par
If $ g^{jn} = hg^{-i}, $ i.e. $ h = g^{i+jn}, $ then the problem is solved. If 
the lists are disjoint, then $ h $ cannot be represented as $ g^{i + jn}, i, 
j\leq n,$. As all powers of $ g $ are thus recorded, the logarithm problem does 
not have a solution.
\end{Proof}

You can use the baby-step-giant-step algorithm to demonstrate that it is much 
more difficult to calculate the discrete logarithm than to calculate the 
discrete exponential function. If the numbers that occur have approximately 1000 
bits in length, then you only need around 2000 multiplications to calculate $ 
g^k $ but around $ 2^{500} \approx 10^{150} $ operations to calculate the 
discrete logarithm using the baby-step-giant-step algorithm. \\
In addition to the baby-step-giant-step algorithm, there are also numerous other 
procedures for calculating the discrete logarithm \cite{4Stinson1995}\index{Stinson 1995}.

### 2007-07-05 script/en/chapters/numbertheory.tex
Update scripts: minimal correction (1.4.10 in menutree)

\item[{\bf 2.}] Select an arbitrary $e \in \{2, \cdots, n-1\}$ such that\footnote{%
                It is recommended by cryptanalytic reasons, but not necessary to
                make RSA work, to select $e$ such that: \\
                $\max(p,q)  <  e <  J(n) - 1$.}: \\
                $e$ is relatively prime to $J(n) = (p-1)*(q-1)$. \\
                We can then ``throw away'' $p$ and $q$.\footnote{%
                The procedure also allows us to select $d$ freely and then 
                calculate $e$. However, this has practical disadvantages.
                We usually want to be able to encrypt messages ``quickly'', which 
                is why we choose a public exponent $e$ such that 
                it has a short bit length compared to the modulus $n$ and 
                as few binary ones as possible (e.g. $2^{16} + 1$).
                So a fast exponentiation is possible when encrypting.                
                We want to select the publicly known $e$ to be an advantageous 
                value that allows the exponential calculation to be performed 
                quickly during encryption. The prime numbers $3, 17$ and $65537$ 
                have proved to be particularly practical for this purpose.
                The most often used number is $65537 = 2^{16}+1$, or in binary: 
                $10\cdots 0\cdots 01$ (this number is prime and therefor relatively
                prime to many other numbers).
                }
\item[{\bf 3.}] Select $d \in \{1, \cdots, n-1\}$ with $e*d \equiv 1  
                {\rm ~(mod~} J(n))$, i.e. $d$ is the multiplicative inverse
                of $e$ modulo $J(n)$\footnote{For reasons of security, 
                $d$ should not be too small.}$^,$\footnote{We start by determining either $d$ or $e$ depending 
                on the implementation.}. We can then ``throw away'' $J(n)$.
\begin{itemize}
    \item[] $\rightarrow (n, e)$ is the public key $P$.
        \item[] $\rightarrow (n, d)$ is the secret key $S$ (only $d$ must be kept secret).
\end{itemize}
\item[{\bf 4.}] For encryption, the message represented as a (binary) number is divided into parts such that each part of 
                the number is less than $n$.
\item[{\bf 5.}] Encryption of the plaintext (or the parts of it) $M \in \{1, \cdots, n-1\}$:
                $$C = E ( (n, e); M ) := M^e {\rm ~(mod~} n).$$
\item[{\bf 6.}] For decryption, the cipher text represented as a binary number is divided into parts such that each part 
                of the number is less than $n$.
\item[{\bf 7.}] Decryption of the cipher text (or the parts of it) $C \in \{1, \cdots, n-1\}$:
                $$M = D ( (n, d); C ) := C^d {\rm ~(mod~} n).$$
\end{itemize}
The numbers $d, e$ and $n$ are usually extremely large (e.~g.\ $d$ and $e$ $300$ bits, $n$ $600$ bits).

@@

Before applying the RSA procedure to a text, we will first demonstrate it
directly using a single number as message\footnote{Using CrypTool\index{CrypTool} you can solve this with the menu 
{\bf Indiv.Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.}.
\begin{itemize}
\item[{\bf 1.}] Let the selected prime numbers be $p=5$ and $q=11$. \\
Thus, $n=55$ and $J(n)=(p-1)*(q-1)=40$.
\item[{\bf 2.}] $e = 7$ (should lie between $11$ and $40$ and must be relatively prime to $40$).
\item[{\bf 3.}] $d = 23$ (since $23*7 \equiv 161 \equiv 1 {\rm ~(mod~} 40)$),
    \begin{itemize}
    \item[] $\rightarrow$ Public key of the recipient:  $(55, 7),$
    \item[] $\rightarrow$ Private key of the recipient: $(55, 23).$
    \end{itemize}
\item[{\bf 4.}] Let the message be the number $M = 2$ (so no division into blocks is required).
\item[{\bf 5.}] Encryption: $C \equiv 2^7 \equiv 18 {\rm ~(mod~}55).$
\item[{\bf 6.}] The cipher text is simply the number $C = 18$ (we therefore do not need to divide it into blocks).
\item[{\bf 7.}] Decryption: 
        $M \equiv 18^{23} \equiv 18^{(1+2+4+16)} \equiv 18*49*36*26 \equiv 2 {\rm ~(mod~}55).$
\end{itemize}

@@

{\bf Key generation (steps 1 to 3)}:\\
{\bf 1.} $p=47, q=79$ $( n= 3,713; ~J(n) = (p-1)*(q-1)=3,588).$ \\
{\bf 2.} $e = 37$ (should lie between $79$ and $3,588$ and must be relatively prime to $3,588$). \\
{\bf 3.} $d=97$ (since $e*d=1{\rm ~mod~}J(n); 37*97 \equiv 3,589 \equiv 1{\rm ~(mod~}3,588) \;$)\footnote{How
to compute $d = 97$ using the {\em extended} gcd algorithm is shown in \hyperlink{Appendix_A}{appendix A of this chapter}}. \\

@@

{\bf 2.} $e=65,537$ (should lie between $509$ and $255,016$ and must be relatively prime to $255,016$)\footnote{$e$ cannot,
therefore, be $2, 127$ or $251$ ($65537 = 2^{16}+1$).\\
In real life, $J(n)$ is not factorised but rather the Euclidean algorithm 
is used for the selected e to guarantee that ${\rm gcd}(d,J(n))=1$.}.\\

@@

This can be accomplished as follows. First, factor $n$ (which is easy because it is so small). Then compute the 
exponent $d$ from $J(n)$, and, finally, decrypt the cipher text. Use the square-and-multiply\index{Square and multiply} algorithm to exponentiate 
modulo $n$.

@@

\newpage
TABLE 4.1\footnote{The numbers of this table can be worked with via Copy and Paste.}: RSA cipher text

>>>>

\item[{\bf 2.}] Select an arbitrary $e \in \{2, \cdots, n-1\}$ such
                that\footnote{%
                \label{foot:Selection-of-e}%
                It is recommended by cryptanalytic reasons, but not
                necessary to make RSA work, to select $e$ such that: \\
                $\max(p,q)  <  e <  J(n) - 1$.}: \\
                $e$ is relatively prime to $J(n) = (p-1)*(q-1)$. \\
                We can then ``throw away'' $p$ and $q$.\footnote{%
                The procedure also allows us to select $d$ freely and then 
                calculate $e$. However, this has practical disadvantages.
                We usually want to be able to encrypt messages ``quickly'',
                which is why we choose a public exponent $e$ such that 
                it has a short bit length compared to the modulus $n$ and 
                as few binary ones as possible (e.g. $2^{16} + 1$).
                So a fast exponentiation is possible when encrypting.                
                We want to select the publicly known $e$ to be an
                advantageous value that allows the exponential calculation
                to be performed quickly during encryption. The prime numbers
                $3, 17$ and $65537$ have proved to be particularly practical
                for this purpose.
                The most often used number is $65537 = 2^{16}+1$, or in
                binary: 
                $10\cdots 0\cdots 01$ (this number is prime and therefore
                relatively prime to many other numbers).
                }
\item[{\bf 3.}] Select $d \in \{1, \cdots, n-1\}$ with $e*d \equiv 1  
                {\rm ~(mod~} J(n))$, i.e. $d$ is the multiplicative inverse
                of $e$ modulo $J(n)$\footnote{For reasons of security, 
                $d$ should not be too small.}$^,$\footnote{We start by determining either $d$ or $e$ depending 
                on the implementation.}. We can then ``throw away'' $J(n)$.
\begin{itemize}
    \item[] $\rightarrow (n, e)$ is the public key $P$.
        \item[] $\rightarrow (n, d)$ is the secret key $S$ (only $d$ must be kept secret).
\end{itemize}
\item[{\bf 4.}] For encryption, the message represented as a (binary) number is divided into parts such that each part of 
                the number is less than $n$.
\item[{\bf 5.}] Encryption of the plaintext (or the parts of it) $M \in \{1, \cdots, n-1\}$:
                $$C = E ( (n, e); M ) := M^e {\rm ~(mod~} n).$$
\item[{\bf 6.}] For decryption, the cipher text represented as a binary number is divided into parts such that each part 
                of the number is less than $n$.
\item[{\bf 7.}] Decryption of the cipher text (or the parts of it) $C \in \{1, \cdots, n-1\}$:
                $$M = D ( (n, d); C ) := C^d {\rm ~(mod~} n).$$
\end{itemize}
The numbers $d, e$ and $n$ are usually extremely large (e.~g.\ $d$ and $e$ $300$ bits, $n$ $600$ bits).

@@

Before applying the RSA procedure to a text, we will first demonstrate it
directly using a single number as message\footnote{Using CrypTool\index{CrypTool} you can solve this with the menu 
{\bf Indiv.Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.}.
\begin{itemize}
\item[{\bf 1.}] Let the selected prime numbers be $p=5$ and $q=11$. \\
Thus, $n=55$ and $J(n)=(p-1)*(q-1)=40$.
\item[{\bf 2.}] $e = 7$ (should\footnote{%
                See footnote \ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $11$ and $39$ and must be relatively prime
                to $40$).
\item[{\bf 3.}] $d = 23$ (since $23*7 \equiv 161 \equiv 1{\rm ~(mod~} 40)$),
    \begin{itemize}
    \item[] $\rightarrow$ Public key of the recipient:  $(55, 7),$
    \item[] $\rightarrow$ Private key of the recipient: $(55, 23).$
    \end{itemize}
\item[{\bf 4.}] Let the message be the number $M = 2$ (so no division into blocks is required).
\item[{\bf 5.}] Encryption: $C \equiv 2^7 \equiv 18 {\rm ~(mod~}55).$
\item[{\bf 6.}] The cipher text is simply the number $C = 18$ (we therefore do not need to divide it into blocks).
\item[{\bf 7.}] Decryption: 
        $M \equiv 18^{23} \equiv 18^{(1+2+4+16)} \equiv 18*49*36*26 \equiv 2 {\rm ~(mod~}55).$
\end{itemize}

@@

{\bf Key generation (steps 1 to 3)}:\\
{\bf 1.} $p=47, q=79$ $( n= 3,713; ~J(n) = (p-1)*(q-1)=3,588).$ \\
{\bf 2.} $e = 37$ (should\footnote{%
                See footnote \ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $79$ and $3,587$ and must be relatively prime
                to $3,588$). \\
{\bf 3.} $d=97$ (since $e*d=1{\rm ~mod~}J(n); 37*97 \equiv 3,589 \equiv 1{\rm ~(mod~}3,588) \;$)\footnote{How
to compute $d = 97$ using the {\em extended} gcd algorithm is shown in \hyperlink{Appendix_A}{appendix A of this chapter}}. \\

@@

{\bf 2.} $e=65,537$ (should\footnote{%
                See footnote \ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $509$ and $255,015$ and must\footnote{%
       $e$ cannot, therefore, be $2, 127$ or $251$
       ($65,537 = 2^{16}+1$) ($255,016 = 2^{3}*127*251$).\\
       In real life, $J(n)$ is not factorised but rather the Euclidean
       algorithm is used for the selected e to guarantee that
       ${\rm gcd}(e,J(n))=1$.
                 } be relatively prime to $255,016$).\\

@@

This can be accomplished as follows. First, factor $n$ (which is easy
because it is so small). Then compute the exponent $d$ from $J(n)$, and,
finally, decrypt the cipher text. Use the square-and-multiply
\index{Square and multiply} algorithm to exponentiate modulo $n$.

@@

% \newpage  be_05.07.07 raus, da sonst vorherige Seite fast leer.
\vspace{12pt}
TABLE 4.1\footnote{The numbers of this table can be worked with via Copy and Paste.}: RSA cipher text
